---
title: "Validation of Bioequivalence Test Performed by BE R package"
author: "Sungpil Han <shan@acp.kr>"
date: "`r Sys.Date()`"
bibliography: ['manual.bib', 'packages.bib']
#bibliography: 'packages.bib'
always_allow_html: yes
header-includes:
  - \usepackage{textcomp}
output: 
  bookdown::pdf_document2: 
    df_print: "kable"
  pdf_document: default
---

```{r include=FALSE}
library(knitr)
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
options(knitr.kable.NA = '')
```

# Introduction

To assess bioequivalence, the 90% confidence interval for the difference in the means of the log-transformed data should be calculated using appropriate methods to the study design. 
The antilogs (exponents) of the confidence limits obtained constitute the 90% confidence interval for the ratio of the geometric means between the T and R products. [@fda;@chow;@hauschke] 
To establish bioequivalence, the calculated confidence interval should fall within a bioequivalence limit, usually 80-125% for the ratio of the product averages. 
For nonreplicated crossover designs, general linear model procedures available in PROC GLM in SAS are preferred, although linear mixed-effects model procedures can also be indicated for analysis. [@fda]

`BE` R package [@R-BE] can analyze bioequivalence study data with industrial strength. 
The current version of `BE` performs bioequivalency tests for several pharmacokinetic parameters of the conventional two-treatment, two-period, two-sequence (2x2) randomized crossover design. 
The statistical model includes factors accounting for the following sources of variation: sequence (SEQ), subjects nested in sequences (SUBJ(SEQ)), period (PRD), and treatment (TRT). 

In this document, the author performed validation of bioequivalence tests performed by `BE` R package as compared to bioequivalence tests performed by PROC GLM or PROC MIXED in SAS.

\pagebreak

# Methods

## Dataset

A simulated dataset of the conventional 2×2 crossover study for this analysis, `BE::NCAResult4BE` is shown in Appendix A. 
The number of subjects in the sequence 'RT' is 17 and that in the sequence 'TR' is 16. (total N=33)
The 4 variables, SUBJ (subject), GRP (group or sequence), PRD (period), and TRT (treatment), and the 3 pharmacokinetic parameters, AUC~last~, C~max~, and T~max~ are presented and there is no missing values. (total 66 observations with 7 variables)

## Bioequivalence tests in R

The required R packages are following. 

```{r setup, message = FALSE}
library(BE)         # install.packages("BE", repos="http://r.acr.kr")
library(dplyr)      # install.packages("dplyr")
library(readr)      # install.packages("readr")
```

`tab_r_be_results` function uses `BE::test2x2()` and returns the 90% confidence interval.

```{r}
tab_r_be_results <- function(parameter){
  BE::test2x2(BE::NCAResult4BE, parameter)[[4]] %>% 
  as.data.frame() %>% 
  mutate(Analysis = 'R: BE package') %>% 
  select(Analysis, `Lower Limit`, `Point Estimate`, `Upper Limit`)
}
```

## Bioequivalence tests in SAS

To run BE analysis, PROC GLM and PROC MIXED in SAS version 9.4 were used.
The SAS program statements include the variables, SEQ (sequence), TRT (treatment), SUBJ (subject), and PRD (period).
LNAUCL (log-transformed AUC~last~) or LNCMAX (log-transformed C~max~) denotes the response measure. 
A part of SAS scripts are shown below and the full SAS scripts are appended in Appendix B.

```r
PROC GLM DATA=BE OUTSTAT=STATRES; /* GLM use only complete subjects. */
CLASS SEQ PRD TRT SUBJ;
MODEL LNAUCL = SEQ SUBJ(SEQ) PRD TRT;
RANDOM SUBJ(SEQ)/TEST;
LSMEANS TRT /PDIFF=CONTROL('R') CL ALPHA=0.1 COV OUT=LSOUT;
```

```r
PROC MIXED DATA=BE; /* MIXED uses all data. */
CLASS SEQ TRT SUBJ PRD;
MODEL LNAUCL = SEQ PRD TRT;
RANDOM SUBJ(SEQ);
ESTIMATE 'T VS R' TRT -1 1 /CL ALPHA=0.1;
ODS OUTPUT ESTIMATES=ESTIM COVPARMS=COVPAR;
```

A function, `tab_sas_proc_results()` reads SAS analysis results exported to Microsoft Excel files (`.xls`) and converted to comma separated version file (`.csv`). It returns a data frame of 90% confidence interval calculated either `PROC GLM` or `PROC MIXED` in SAS.

```{r}
tab_sas_proc_results <- function(filename, skip_no, analysis_name){
  read_lines(filename, skip = skip_no, n_max = 2) %>% 
  paste(collapse='\n') %>% read_csv() %>% 
  mutate(Analysis = analysis_name) %>% 
  select(Analysis, `Lower Limit` = LL, `Point Estimate` = PE, `Upper Limit` = UL)
}
```

\pagebreak

# Results

## AUC~last~

Comparison of 90% confidence interval for the ratio of the geometric means of AUC~last~ between the T and R products is shown in Table \@ref(tab:tabauclast).

```{r}
AUClast_R_BE <- tab_r_be_results("AUClast")

AUClast_proc_glm <- tab_sas_proc_results('sas/SAS_results_AUClast.csv', 
                                         skip = 206, 'SAS: PROC GLM')
AUClast_proc_mixed <- tab_sas_proc_results('sas/SAS_results_AUClast.csv', 
                                           skip = 278, 'SAS: PROC MIXED')

# Combine all analyses of AUClast
AUClast_all_analyses <- bind_rows(AUClast_R_BE, AUClast_proc_glm, AUClast_proc_mixed)
```

```{r tabauclast, echo = FALSE}
kable(AUClast_all_analyses,
      longtable = TRUE, booktabs = TRUE, digits = 5,
      caption = 'Comparison of 90\\% confidence interval for the ratio of the geometric means of AUClast')
```

## C~max~

Comparison of 90% confidence interval for the ratio of the geometric means of AUC~last~ between the T and R products is shown in Table \@ref(tab:tabcmax).

```{r }
Cmax_R_BE <- tab_r_be_results("Cmax")

Cmax_proc_glm <- tab_sas_proc_results('sas/SAS_results_Cmax.csv', 
                                      skip = 206, 'SAS: PROC GLM')
Cmax_proc_mixed <- tab_sas_proc_results('sas/SAS_results_Cmax.csv', 
                                        skip = 278, 'SAS: PROC MIXED')

# Combine all analyses of Cmax
Cmax_all_analyses <- bind_rows(Cmax_R_BE, Cmax_proc_glm, Cmax_proc_mixed)
```

```{r, tabcmax, echo = FALSE}
kable(Cmax_all_analyses, longtable = TRUE, booktabs = TRUE, digits = 5, escape = TRUE,
      caption = 'Comparison of 90\\% confidence interval for the ratio of the geometric means of Cmax')
```

\pagebreak

# Conclusion 

*There is no discrepancy* between bioequivalence tests performed by `BE` R package and those performed by PROC GLM or PROC MIXED in SAS.
We also performed multiple analyses with the actual clinical trial datasets and have found no differences (data not shown: confidential).

Bioequivalence tests performed by the open-source `BE` R package for the conventional two-treatment, two-period, two-sequence (2x2) randomized crossover design can be **qualified and validated** enough to acquire the identical results of the commercial statistical software, SAS.

*Please report issues regarding validation of the R package to <https://github.com/asancpt/BEreport/issues>.*

-----

**Affiliation**:  
Sungpil Han M.D/Ph.D  
Resident,  
Department of Clinical Pharmacology and Therapeutics,  
Asan Medical Center, University of Ulsan,  
Seoul 05505, Republic of Korea  
E-mail: shan@acp.kr  
URL: www.github.com/shanmdphd  

\pagebreak

# (APPENDIX) Appendix {-}

# Raw data

The concentration-time curves are ploted in Figure \@ref(fig:conctime) and the result of noncomparmental analysis is presented in Table \@ref(tab:rawdata).

```{r conctime, echo = FALSE, fig.cap = 'Concentration-time curves'}
knitr::include_graphics('assets/conc-time.pdf')
```

```{r rawdata, echo = FALSE}
kable(BE::NCAResult4BE, 
      caption = 'The raw data used for analysis in R and SAS', 
      longtable = TRUE, booktabs = TRUE)
```

# SAS Scripts and results

To run these scripts, the dataset `BE::NCAResult4BE` should be exported from R by `write.csv()`.

## AUC~last~

```{r, code = readLines('sas/sas-be-auclast.sas'), eval = FALSE}
```

```{r echo = FALSE}
read_csv('sas/SAS_results_AUClast.csv', skip=143, n_max = 4) %>% 
  select(1:6) %>% 
  knitr::kable(longtable = TRUE, booktabs = TRUE, 
               caption = 'Table of analysis of variance for log-transformed AUClast (PROC GLM)') 
```

## C~max~

```{r, code = readLines('sas/sas-be-cmax.sas'), eval = FALSE}
```

```{r echo = FALSE}
read_csv('sas/SAS_results_Cmax.csv', skip=143, n_max = 4) %>% 
  select(1:6) %>% 
  knitr::kable(longtable = TRUE, booktabs = TRUE, 
               caption = 'Table of analysis of variance for log-transformed Cmax (PROC GLM)') 
```

# Session Information

```{r}
devtools::session_info()
```

\pagebreak

# References {-}

```{r include = FALSE}
library(BE)
knitr::write_bib(file = 'packages.bib')
```

```{r, eval = FALSE, include = FALSE}
BE::test2x2(BE::NCAResult4BE, "Cmax")

anova_be <- BE::test2x2(NCAResult4BE, "Cmax")[[1]] %>% 
  as.data.frame() %>% 
  mutate(Source = row.names(.)) %>% 
  select(Source, SS, DF, F) %>% 
  gather(parameter, value, -Source) %>% 
  mutate_at(vars(value), funs(round), 5)
  
anova_sas <- read_lines('sas/SAS_results_Cmax.csv', skip = 181, n_max = 10) %>% 
  paste(collapse = '\n') %>% 
  read_csv() %>% 
  rename(Name=2, Source=3, Type=4) %>% 
  filter(Type %in% c('SS3', 'ERROR')) %>% 
  select(Source, SS, DF, F) %>% 
  mutate(Source = case_when(
    Source == 'SEQ' ~ 'GROUP',
    Source == 'SUBJ(SEQ)' ~ 'SUBJECT(GROUP)',
    Source == 'PRD' ~ 'PERIOD',
    Source == 'TRT' ~ 'DRUG',
    TRUE ~ Source
    )) %>% 
  gather(parameter, value, -Source) %>% 
  mutate_at(vars(value), ~ as.numeric(.x) %>% round(5)) %>% 
  print()

left_join(anova_be, anova_sas, by = c('Source', 'parameter'))
```

